---
title: "Real-world Data Analysis Using R: Subset of Framingham Data"
author: "Mina Peyton"
date: "2024-09-20"
output: html_document
---


```{r}
# if needed, install required packages for this script
# install.packages(c("tidyverse", "readr", "psych", "car", "moments", "ggplot2",
# "dunn.test", "rstatix")
```


```{r}
# attach libraries that are need to run the script
library(tidyverse)
library(readr)
library(psych)
library(car)
library(moments)
library(ggplot2)
library(dunn.test)
library(rstatix)
```


Read in framingham.csv data file and store it as a data frame object name "df".  
The "df" object is now listed in the "Environment" 
df with 4240 obs of 16 variables.  

You can click on "df" to open in a new tab to view the data.  

```{r}
df = read_csv("framingham.csv")
str(df)

```

# Objective 1:    
Determine if there is a difference in average BMI for individuals with high cholesterol 
(cholesterol ≥ 240 mm/L) compared to individuals without high cholesterol (i.e., 
with normal/borderline cholesterol levels).  

Total cholesterol: Less than 200 mg/dL is normal, 200–239 mg/dL is borderline high, 
and 240 mg/dL or higher is high  
(https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel)  

What is your research question?  
Is there a difference in average BMI in individuals with high cholesterol compared
to those that do not have high cholesterol?  

What is the study design?  
What is the population of interest?  

Create two new variables from the total cholesterol level info above.  
Chol_group to define the three levels above (normal, borderline, high)  
Chol_bin to define two levels as binary (0 - do not have high cholesterol, 
1 - have high cholesterol)

```{r}
df <- df %>%
  mutate(Chol_group = case_when(
    totChol <= 200 ~ "normal",
    totChol > 200 & totChol <= 239 ~ "borderline",
    totChol >= 240 ~ "high")) %>%
  mutate(Chol_bin = ifelse(Chol_group == "high", 1, 0))

str(df)
```


Set the new variables as factors

```{r}
df <- df %>%
  mutate(Chol_group = as.factor(case_when(
    totChol <= 200 ~ "normal",
    totChol > 200 & totChol <= 239 ~ "borderline",
    totChol >= 240 ~ "high"))) %>%
  mutate(Chol_bin = as.factor(ifelse(Chol_group == "high", 1, 0)))

str(df)
```

Re-order the levels of the Chol_group

```{r}
df$Chol_group = factor(df$Chol_group, levels = c("normal", "borderline", "high"))
str(df)

```

Exploratory data analysis: Explore the variables of interest

```{r}
table(df$Chol_group)
```

```{r}
table(df$Chol_bin)
```

Find and remove NA values

```{r}
sum(is.na(df$BMI)) # 19
```

```{r}
sum(is.na(df$totChol)) # 50
```

```{r}
remove = c(which(is.na(df$BMI)),which(is.na(df$totChol)))
```

```{r}
df = df[-remove,] # [rows, columns]
dim(df)
```

Create graphical summaries that visualizes BMI for individuals with high cholesterol 
compared to individuals without high cholesterol

```{r}
hist(df$BMI[df$Chol_bin == 0], breaks = 20, xlim = c(10, 60), ylim = c(0,500))
```

```{r}
hist(df$BMI[df$Chol_bin ==1], breaks = 20, xlim = c(15, 45), ylim = c(0,250))
```

```{r}
boxplot(df$BMI ~ df$Chol_bin, data = df)
```

Calculate summary statistics for BMI for both group (with and without high cholesterol)

```{r}
#library(psych)
describeBy(df$BMI, df$Chol_bin)
```

```{r}
sum_stats = df %>% group_by(Chol_bin) %>%
  summarise(
    n = n(),
    mean = mean(BMI),
    sd = sd(BMI),
    se = sd/sqrt(n),
    median = median(BMI))
```

Determine statistical test that would be most appropriate for answering the research question: 
one-sample t-test, paired t-test, two-sample t-test (assuming unequal variances), 
or two-sample t-test (assuming equal variances).  

Check assumptions for your test.  

```{r}
# Check normality
shapiro.test(df$BMI[df$Chol_bin==0])
shapiro.test(df$BMI[df$Chol_bin==1])
# Sensitive to large sample sizes
```

```{r}
qqnorm(df$BMI[df$Chol_bin==0], main = "Low Cholesterol Group")
qqline(df$BMI)
```

```{r}
qqnorm(df$BMI[df$Chol_bin==1], main = "High Cholesterol Group")
qqline(df$BMI)
```

```{r}
# check equal variance 

# Levene's Test
leveneTest(BMI ~ Chol_bin, data = df)
# null hypothesis = variances of the groups are equal
# alternative hypothesis = variances of the groups are unequal
# p = 0.21, fail to reject the null hypothesis, variance of the groups are equal
```

Statistical Inference: Carry out the hypothesis test. 

- What are the hypotheses?  

null hypothesis = no difference in mean BMI between both groups  
alternative hypothesis = there is a difference in mean BMI between both groups   
  
- Your p-value?  
- Make a conclusion using α = 0.05  
  
```{r}
# Two-sample t-test with equal variance
t.test(df$BMI~df$Chol_bin, alternative="two.sided", var.equal=TRUE)
# 9.201e-12
```

```{r}
# Wilcoxon Rank-Sum Test
wilcox.test(df$BMI[df$Chol_bin==0], df$BMI[df$Chol_bin==1])
# p-value = 2.957e-14
```

Conclusion: Provide an answer to your research question.  
There is a significant difference (p-value = 9.201e-12) in mean BMI between individuals with
high cholesterol compared to those without high cholesterol. The high cholesterol group
had higher BMI (26.3 +- 3.96) compared to non-high cholesterol group (25.4 +- 4.13).  

Data visualization  
(https://ggplot2-book.org/)  

*statistically different from non-high cholesterol  
```{r}
p <- ggplot(sum_stats, aes(x = Chol_bin, y = mean, fill = Chol_bin)) +
  geom_bar(stat = "identity", color = "black") +
  geom_errorbar(aes(ymin = mean, ymax = mean + sd), width = 0.2) +
  scale_x_discrete(labels = c("0" = "Non-high Cholesterol", "1" = "High Cholesterol")) +
  theme_minimal() +
  labs(title = "Mean BMI by Cholesterol Group",
       x = "Cholesterol Groups",
       y = "BMI") +
  theme(legend.position = "none")  # Hide the legend

# Add annotations
p + annotate("text", x = 2, y = (sum_stats$mean[1] + sum_stats$sd[1]) + 1.5, label = "*", size = 8, color = "black")  

```

# Pratice Exercise    
Determine if there is a difference in totChol between males and females  
males = 1  
females = 0  

```{r}

```


# Objective 2:  
Determine if there is a difference in BMI between the Chol_group (i.e,
normal, borderline, and high cholesterol groups)?

```{r}
sum_stats = df %>% group_by(Chol_group) %>%
  summarise(
    n = n(),
    mean = mean(BMI),
    sd = sd(BMI),
    se = sd/sqrt(n),
    median = median(BMI))

describeBy(df$BMI, df$Chol_group)
```

```{r}
qqnorm(df$BMI[df$Chol_group== "normal"], main = "Normal Cholesterol Group")
qqline(df$BMI)
```

```{r}
qqnorm(df$BMI[df$Chol_group== "borderline"], main = "Borderline Cholesterol Group")
qqline(df$BMI)
```

```{r}
qqnorm(df$BMI[df$Chol_group== "high"], main = "High Cholesterol Group")
qqline(df$BMI)
```

```{r}
# Equal variance: Brown-Forsythe Test 
leveneTest(BMI ~ Chol_group, data = df)
# pvalue =0.4385
```

Statistical inference: One-way ANOVA (Analysis of Variance)
```{r}
# Perform one-way ANOVA
anova_model <- aov(BMI ~ Chol_group, data = df)

# View the ANOVA table
summary(anova_model)
```


```{r}
# Check normality of residuals
qqnorm(residuals(anova_model))
qqline(residuals(anova_model))
```


```{r}
# Perform Tukey's Honest Significant Differences test
# post-hoc test used after performing an ANOVA to find out which specific group means 
# are significantly different from each other. 
# It compares all possible pairs of means and adjusts for multiple comparisons, 
# to control the family-wise error rate
posthoc <- TukeyHSD(anova_model)
posthoc
```
There is a significant difference (p = 7.42e-15) in mean BMI between cholesterol groups.
Tukey's HSD post-hoc analysis revealed significant differences between borderline (p = 4.84e-05)
and high (p = 0.00e+00) cholesterol groups compared to normal, and a significant difference in 
mean BMI between high compared to borderline cholesterol groups (p = 1.65e-04). Overall mean BMI
for normal, borderline, and high cholesterol groups were 24.9 +- 4.18, 25.7 +- 4.07, and 26.3 +- 3.96,
respectively.  

Data visualization  
 *significantly different from normal  
 # significantly different from borderline

```{r}
# Create the boxplot
p <- ggplot(sum_stats, aes(x = Chol_group, y = mean, fill = Chol_group)) +
  geom_bar(stat = "identity", color = "black") +
  geom_errorbar(aes(ymin = mean, ymax = mean + sd), width = 0.2) +
  theme_minimal() +
  labs(title = "Mean BMI by Cholesterol Group",
       x = "Cholesterol Groups",
       y = "BMI") +
  theme(legend.position = "none")  # Hide the legend

# Add annotations
# Assuming these positions are appropriate for your plot
p + 
  annotate("text", x = 2, y = (sum_stats$mean[1] + sum_stats$sd[1]) + 1, label = "*", size = 8, color = "black") +  # Normal vs. Borderline
  annotate("text", x = 2.9, y = (sum_stats$mean[2] + sum_stats$sd[2]) + 1, label = "*", size = 8, color = "black") +  # Normal vs. High
  annotate("text", x = 3.1, y = (sum_stats$mean[3] + sum_stats$sd[3]) + 1, label = "#", size = 5, color = "black")  # Borderline vs. High
```

If the normality assumption is violated: Kruskal-Wallis test
```{r}
kruskal.test(BMI ~ Chol_group, data = df)
```

Non-parametric post-hoc comparisons after a Kruskal-Wallis test: Dunn's test or the pairwise Wilcoxon rank-sum test  

Dunn's Test  
```{r}
dunn.test(df$BMI, df$Chol_group, method = "bonferroni") # or "hs" for Holm-Sidak adjustment
```

Pairwise Wilcoxon rank-sum test  
```{r}
pairwise.wilcox.test(df$BMI, df$Chol_group, p.adjust.method = "bonferroni")
```

If equal variance assumption is violated: Welch’s ANOVA
```{r}
oneway.test(BMI ~ Chol_group, data = df, var.equal = FALSE)
```
Pairwise Comparisons with Games-Howell Test

```{r}
games_howell_test(BMI ~ Chol_group, data = df)

```




































































































