---
title: "Part 1_Intro to R_Data wrangling"
author: "Yuyan Yi"
date: "2024-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

1. R install
2. Operators
3. library install, load
4. dataset import


## 1. Installing required software and background

The course requires an installation of R version 4.3.3 and Rstudio.  You may install R and Rstudio directly from the following web sites:

* [R install](https://cran.r-project.org/)
* [R studio](https://www.rstudio.com/products/rstudio/)

Additional information on manual installation can be found [HERE](https://rstudio-education.github.io/hopr/starting.html#how-to-download-and-install-r)

Alternatively, if within the NIAID enterprise, you may install using self-service:

* [Windows Software Center](https://inside.niaid.nih.gov/it-equipment/installing-scientific-software-using-software-center)
* [Mac Self Service](https://inside.niaid.nih.gov/it-equipment/installing-scientific-software-mac-self-service)

* The difference between R and RStudio:
R is a programming language and software environment for statistical computing and graphics, RStudio is an integrated development environment (IDE) that provides a user-friendly interface for working with R. RStudio is designed to enhance the R experience by providing tools and features that make it easier to write, test, and debug R code, as well as manage projects and data.


##### 1.1 Create workspace
* create/close project
* panel layout (Tools--Global Options--Panel Layout)
* save script

##### 1.2 Set working directory  
* set work directory: setwd("file_path") or setwd(file.choose()) to manually select a path   
* get work directory: getwd()  

```{r eval=FALSE}
setwd("~/OneDrive - National Institutes of Health/Documents/BCBB training course/2024_R")
getwd()
```

## 2 Operator
* see the full list from [manual:3.1.4 Operators](https://cran.r-project.org/doc/manuals/r-release/R-lang.pdf)  

| Operator         | Description                                    |
|------------------|------------------------------------------------|
| `<-`             | Left assignment, binary                        |
| `$`              | List subset, binary                            |
| `%in%`           | Matching operator, binary (in model formulae: nesting)  |
| `<`              | Less than, binary                              |
| `>`              | Greater than, binary                           |
| `==`             | Equal to, binary                               |
| `>=`             | Greater than or equal to, binary               |
| `<=`             | Less than or equal to, binary                  |
| `&`              | And, binary, vectorized                        |
| `&&`             | And, binary, not vectorized                    |
| `|`              | Or, binary, vectorized                         |
| `||`             | Or, binary, not vectorized                     |

```{r echo=TRUE}
#create variable
a = 2
a <- 2

#logical 
a < - 2
a == 2
a <= 2

#arithmetic expressions
x <- 2 * a
y1 <- 2 * (a + log(x)) #default natural logarithm
y2 <- 2 * (a + log(x, base = exp(1))) #specify base
y3 <- 2 * (a + log10(x))
y4 <- 2 * (a + log2(x))

#round numbers
y <- round(y1,4)

#create numerical vector
x <- rnorm(10) #random generate a vector that length = 10, standard normal distribution
x[2] #extract a value from vector based on the index
x[2:7] #subset the vector

#create charactor vector
x <- c("apple","pear","peach","watermelon","banana")
x[2]
x[-1] #remove an element from vector based on the index
a = "apple"
a %in% x #check if element is in a vector

#create factor vector
risk <- c(rep("high",20), rep("medium", 30),rep("low",10))
risk1 <- factor(risk)
risk2 <- factor(risk, levels = c("high","medium","low"))

#create list,which could contains multiple different type of elements
L = list()
L[[1]] = "apple"
L[[2]] = c(1:5)
L[[3]] = TRUE
L[[4]] = matrix(rnorm(10),2,5)

#assign the name to each element
names(L) = c("L_Character","L_Number","L_Logical","L_Matrix") 
#extract an element from list by its name
L$L_Number 
#extract an element from list by its index
L[[2]] 

#create matrix
A <- matrix(data = runif(24, min= -1, max = 1), nrow = 4, ncol = 6)
A[2,5] #extract a value from matrix based on the row and column index
A[1:2,] #extract first two rows from matrix
A[,-c(1,3)] #remove 1st, 3rd columns from matrix

#create data frame
D <- data.frame(age = c(2,1,4,3,5), 
                animal = c("dog","cat","sheep","panda","tiger"), 
                ifmale = c(T,F,T,F,T), 
                weight = c(5.2,2.2,30.5,124.1,90.3))
row.names(D) = c("sample1","sample2","sample3","sample4","sample5") #assign row name 
D$animal #subset one variable
D[D$ifmale==T,] #subset rows with condition
D[D$ifmale==T & D$weight>10,] #subset rows with multiple conditions
D[D$ifmale!=T | D$age<=2,] #subset rows with multiple conditions


#remove objects from current environment
rm(A)
remove(A)
A
```


## 3. Installing R packages

Package quality varies and some important considerations on whether a particular package is trustworthy include some of the following considerations shown below.  A great example of a trustworthy package is [ggplot2](https://ggplot2.tidyverse.org/).  Please check out the package website and see if you can find examples of the following.

* Status of development life cycle showing a robust set of testing used to assess package functionality
* Amount of code used in the package that is covered by various tests
* Sharing of the package on CRAN which include a set of standard tests to assess package functionality
* Detailed documentation on what the package does, its functions, and how to use it
* A strong user base (we will check this by installing and using a package called cranlogs below)


##### 3.1 Now let's install our first R package to get some statistics on the number of downloads a particular R package has:

```{r eval=FALSE}
## The install.packages function is used to install new R packages via command line
## We will use it to install cranlogs that can tell us about how many times an R package is downloaded
install.packages("cranlogs")
#install packacges once
install.packages(c("readr", "dplyr", "Stat2Data","ggplot2")) # install multiple packages

#alternative way to install packages
#"Packages" --> "Install" --> e.g."BiocManager"

#install from website
#e.g. Bioconductor
#Bioconductor is a free, open source and open development software project for the analysis and comprehension of genomic data generated by wet lab experiments in molecular biology.


#Install a package from Github
#GitHub is a repository useful for all software development and data analysis, including R packages. It makes sharing your package easy. You can read more about GitHub here: Git and GitHub, by Hadley Wickham.

#To install a package from GitHub, the R package devtools (by Hadley Wickham) can be used. You should first install devtools if you don’t have it installed on your computer.

#For example, the following R code installs the latest version of survminer R package developed by A. Kassambara (https://github.com/kassambara/survminer).
#rare case
install.packages("devtools")
devtools::install_github("kassambara/survminer")
library(survminer)

#load packages everytime when you run project
library(readr)
library(readxl)
library(dplyr)
library(Stat2Data)
library(ggplot2)
library(MASS)
library(BiocManager)
library(dplyr)

```

##### 3.2 Let's use `cranlogs` package to determine how many downloads `ggplot2` package has

```{r}
## To load an R package, we use the library function
library(cranlogs)

## Lets see how many downloads ggplot2 had in 2021
ggplot2_dls <- cran_downloads(packages = "ggplot2", from = "2023-01-01", to = "2023-12-31")

# We can see that the average daily downloads of ggplot2 is over 80k!
mean(ggplot2_dls$count)

```

##### 3.3  Check version of R and loaded packages

```{r}
sessionInfo()
```

##### 3.4 help information

R includes a series of functions for providing the help documentation and other such information concerning packages and their related functions.

There are four commonly used way to search for help:

The `?` and `??` are short hands for help() and help.search() respectively.

Use the first function when you know exactly what to search for, and the second function when you know approximately what to search for.

```{r}
#Documentation Shortcuts
?print
help("print")

#Search the Help System
??print
help.search("print") 
```

## 4. Dataset import and export

*click:
Step1: "Import Dataset"
Step2: "From Text(base)...",for txt and csv files, doesn't work for excel file
       "From Text(readr)...",for txt and csv files, doesn't work for excel file, provide code and preview of imported dataframe (Recommend)
       "From Excel...",for excel file, provide code and preview of imported dataframe (Recommend)
       
*load corresponding package for different type of data files.
*choose appropriate function and specify the type of variable in the dataset.

```{r echo=TRUE}
library(readr)
#text file
framingham.txt <- read.delim("framingham.txt")
framingham.txt <- read_delim("framingham.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE)
#csv file
framingham.csv <- read.csv("framingham.csv")

#excel file
library(readxl)
framingham.xlsx <- read_excel("framingham.xlsx")

#export file
write.csv(x = framingham.csv, file = "framingham_test.csv", row.names = F) # If you do not specify row.names = F then the row names will be written out, there are other function inputs you can look up using help(write.csv)
```



## 5. Basic data manipulations

```{r echo=TRUE}
#dimension of dataframe
dim(framingham.csv)
#basic summary of variables
summary(framingham.csv)
#see all variable type
sapply(framingham.csv, class)
#displays the internal structure of an object such as an array, list, matrix, factor, or data frame
str(framingham.csv)
#first six (default) rows of the data
head(framingham.csv)
head(framingham.csv,n = 10)

#check missing value
is.na(framingham.csv)
#count the number of missing values
sum(is.na(framingham.csv)) 
#count the number of missing value for each column
colSums(is.na(framingham.csv)) 
#remove rows/observations that contains missing value
framingham = na.omit(framingham.csv)
dim(framingham)
summary(framingham)

#focus on one variable
education <- framingham$education
mean(education) # average
sd(education) # standard deviation
sum(education) # summation

unique(education) # return all unique value in factor
length(unique(education)) # return the number of unique value in factor

#convert to factor/categorical variable
summary(education) #summary of numerical variable
education = factor(education)
summary(education) #summary of categorical variable
table(education) # same as summary() for categorical variable

#subset dataframe based on condition
#subset column
# select three columns: education, BMI, male
select <- framingham[,c("education", "BMI", "male")]

#subset row
#method 1
framingham.male = framingham[framingham$male==1,]
framingham.male.obesity= framingham[framingham$male==1 & framingham$BMI >=30,]

#method 2
framingham.male = subset(framingham, male == 1)

#arrange  
# Order the data by ascending age and education
head(arrange(framingham, age))
head(arrange(framingham, age, education))

# Order age by descending
head(arrange(framingham, desc(age)))

#create new variable
framingham$gender <- ifelse(framingham$male==1, "Male", "Female"); gender = framingham$gender
summary(gender)
gender = factor(gender, levels = c("Male", "Female")) #convert to factor variable
summary(gender)

#merge dataframe (Adding Rows)
framingham.male = subset(framingham, male == 1)
framingham.female = subset(framingham, male == 0)
framingham.mergerow = rbind(framingham.male,framingham.female)

#merge dataframe (Adding Columns)
framingham$id = paste0("Sample_",c(1:nrow(framingham)))
framingham.1 = framingham[,c(1:5,18)] #subset 1~5 variables with sample_id
framingham.2 = framingham[,c(6:10,18)] #subset 6~10 variables with sample_id
framingham.mergecol <- merge(framingham.2,framingham.2,by="id")

```


## 6. Data Structure conversion

### 6.1 Dataset with repeated measures
```{r echo=TRUE}

olddata_wide <- read.table(header=TRUE, text='
 subject sex control cond1 cond2
       1   M     7.9  12.3  10.7
       2   F     6.3  10.6  11.1
       3   F     9.5  13.1  13.8
       4   M    11.5  13.4  12.9
')
# Make sure the subject column is a factor
olddata_wide$subject <- factor(olddata_wide$subject)

#from wide to long
#method 1
library(tidyr)
data_long1 <- gather(olddata_wide, condition, measurement, control:cond2, factor_key=TRUE)
data_long1 <- gather(olddata_wide, condition, measurement, control, cond1, cond2)

#method 2
library(reshape2)
# Specify id.vars: the variables to keep but not split apart on
data_long2 <- melt(olddata_wide, id.vars=c("subject", "sex"), variable.name="condition", value.name="measurement")

#from long to wide
olddata_long <- read.table(header=TRUE, text='
 subject sex condition measurement
       1   M   control         7.9
       1   M     cond1        12.3
       1   M     cond2        10.7
       2   F   control         6.3
       2   F     cond1        10.6
       2   F     cond2        11.1
       3   F   control         9.5
       3   F     cond1        13.1
       3   F     cond2        13.8
       4   M   control        11.5
       4   M     cond1        13.4
       4   M     cond2        12.9
')
# Make sure the subject column is a factor
olddata_long$subject <- factor(olddata_long$subject)
#method 1
data_wide1 <- spread(olddata_long, condition, measurement)

#method 2
data_wide2 <- dcast(olddata_long, subject + sex ~ condition, value.var="measurement")

```


## 7. Execute externel R script

##### 7.1  run and save objects from externel R script
```{r}
source("Hello_world.R")
```

##### 7.2  run and save functions defined from externel R script
```{r}
source("functions.R")

#add function with return

#find mode value from vector education
mode(education)
table(education)

#calculate circle_area with radius = 5
r = 5
circle_area(r)
#calculate circumference with radius = 5
circumference(r)
#print the result in a sentence
print(paste0("For circle with r = ", r, ", the area of circle is ", round(circle_area(r),4), ", there circumference of circle is ",round(circumference(r),4), "." ))

#check sign of value
check(10)

```


## 8. Exercise questions

#### 8.0 Check the current path

```{r echo=TRUE}
getwd()
setwd("~/OneDrive - National Institutes of Health/Documents/BCBB training course/2024_R")
getwd()
```


#### 8.1 Import Dataset "BirdNest.csv"

```{r echo=TRUE}
library(readr)
BirdNest <- read_csv("BirdNest.csv")
```

#### 8.2 Check dimension and column names of dataset

```{r echo=TRUE}
dim(BirdNest)
colnames(BirdNest)
```

#### 8.3 Data preprocessing

```{r echo=TRUE}
#remove missing value
BirdNest <- na.omit(BirdNest)
#print first 10 rows of dataset
head(BirdNest,n=10)
#basic summary of variables
summary(BirdNest)
#see all variable type
sapply(BirdNest, class)
#add a new column "measure1" with random normal distribution mean = 10, standard deviation = 0.1
BirdNest$measure1 = rnorm(n=nrow(BirdNest),10,0.1)
#add a new column "measure2" with random uniform distribution a = -1, standard deviation = 1
BirdNest$measure2 = runif(n=nrow(BirdNest),-1,1)
#add a new column "measure3" with random poisson distribution lambda = 5
BirdNest$measure3 = rpois(n=nrow(BirdNest),5)

```

#### 8.4 Data Exploration

```{r echo=TRUE}

#return all unique value with frequency in Location variable of BirdNest
table(BirdNest$Location) 
#return all unique value in factor
unique(BirdNest$Location) 
#return the number of unique value in factor
length(unique(BirdNest$Location)) 

```



## Next steps

Today you have learned a lot about how to manipulate and work with data using the R programming language.  Please feel free to practice what we learned today using "BirdNest.csv" (attached) or your own dataset.

Hopefully this will help you get started with analyzing your own data!  Nonetheless, this is just the beginning of your journey into data science with R.  Below you will see some recommended resources to follow up with to continue learning.

* [Intro to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf)
* [Coursera Data Science Specialization](https://www.coursera.org/specializations/jhu-data-science)
* [R for Data Science Book](https://r4ds.had.co.nz/index.html)
* List and resources of useful R packages:
  + [Rstudio's recommendations](https://support.rstudio.com/hc/en-us/articles/201057987-Quick-list-of-useful-R-packages)
  + [Computerworld list](https://www.computerworld.com/article/2921176/great-r-packages-for-data-import-wrangling-visualization.html)
  + [Awesome R GitHub list](https://github.com/qinwf/awesome-R)
  + [CRAN Package Topics](https://cran.r-project.org/web/views/)
  + [Bioconductor for bioinformatics R packages](https://www.bioconductor.org/)

In the second part, we will introduce some classical and fancy approaches in R for data visualization using packages ggplot2, etc.


